### File: index.qmd
### Author: Armand Bashige
### Portfolio: Data Visualization Portfolio — Post 01_my-first-code
### Purpose: Visualize median household income across Hennepin County zip codes
###          using a choropleth map to reveal geographic income disparities
### Data: Zip_Codes shapefile (local) + zipcodeR package (Census-based income data)
### Last updated: 2025-03-22

---
title: "Median Household Income in Hennepin County"
author: "Armand Bashige"
date: "2025-03-22"
categories: [data visualization, maps]
---

## Overview

This visualization explores the distribution of median household income across zip codes in Hennepin County, Minnesota. Using a choropleth map, we can see how income levels vary geographically throughout the county.
```{r}
#| echo: false
#| message: false
#| warning: false
#| output: false

# Load required libraries for spatial data handling and mapping
library(tidyverse)
library(sf) # For reading and working with shapefiles
library(zipcodeR) # Provides Census-based zip code demographic data
library(tmap) # For creating thematic maps

### Load and prepare spatial data ---------------------------------------------

# Read the Hennepin County zip code shapefile
zips <- st_read(dsn = "C:/Users/bashi/OneDrive/Documents/DST490/2025sem2-Bashigea/day11/shapefiles/Zip_Codes", layer = "Zip_Codes")

# Pull all zip codes in Hennepin County, MN from the zipcodeR Census data
hennepin_zips <- search_county("Hennepin","MN")

# Keep only the columns needed for this visualization
small_hennepin_zips <- hennepin_zips |>
  select(zipcode, median_household_income, post_office_city)

### Join income data to spatial polygons --------------------------------------

# Match income data to zip code boundaries by zip code identifier
# st_make_valid() fixes any geometry errors in the shapefile before plotting
zips_income <- zips |>
  left_join(small_hennepin_zips, by=c("NAME_TXT"="zipcode")) |>
  st_make_valid()
```

## Visualization

```{r}
#| echo: false
# Choropleth map: zip code polygons shaded by median household income
# Darker = higher income (tmap default color scale)
tm_shape(zips_income) + 
  tm_polygons("median_household_income")
```

## Key Takeaways

The map reveals notable income disparities across Hennepin County. Western suburbs tend to show higher median household incomes, while areas closer to downtown Minneapolis show more variation. This type of geographic visualization helps highlight spatial patterns in economic data that might not be obvious from tables or summary statistics alone.