---
title: "YOUR FILE NAME HERE"
author: "YOUR NAME HERE"
format:
  html:
    embed-resources: true
editor: source
---

```{r, setup}
knitr::opts_chunk$set(echo=TRUE)
library(mdsr)
library(tidyverse)
#  tidy=FALSE,     # display code as typed
#  size="small")   # slightly smaller font for code
```

# Section titles are here (you need one hashtags)

## Subsections have 2 hashtags

### Subsubsections have 1 hashtag

-   Italics: wrap your phrase with one star (\*): I *love* DST 490!
-   Bold: wrap your phrase with two stars (\*\*): The **class** will be awesome fun!

You can wrap `R` commands in "chunks", using the syntax below

**USE GLIMPSE TO PREVIEW DATASETS.**

Do this:

```{r}
glimpse(mpg)
```

Not that:

```{r}
mpg
```

## Lists and check boxes
For each statement below, if it is true, replace the empty box ("[ ]", rendered as &#x2610;) with a check mark ("[X]", rendered as &#x2611;) in the Quarto file before rendering.

Choose one:

- [ ] `R` is awesome.
- [X] Why didn't you check the previous statement? `R` is still awesome.

## Recreate this plot!

I will show a plot on the google slides. This plot was created using the data verbs in class. Working with your partner, see if you can construct a series of pipes to recreate the plot. Here is some code to get you started:

```{r}
# Location of the excel file:
url <- 'https://raw.githubusercontent.com/jmzobitz/HNDSDatasets/refs/heads/main/data-raw/unemployment_data.csv'

input_data <- read_csv(file = url)

glimpse(input_data)

```
