```{r}
#| echo: false
#| include: false
### File: index.qmd
### Author: Armand Bashige
### Portfolio: Data Visualization Portfolio — Reworked SNAP Enrollment Gap
### Purpose: Reworked visualization showing where SNAP/MFIP enrollment gaps exist
###          across Hennepin County census tracts by poverty quartile (2023-2025)
### Data: HennepinCounty_SNAP_MFIP_Estimates_Tract_RevA.csv (private)
### Last updated: 2026-04-25
```
---
title: "Where Is the SNAP Enrollment Gap?"
author: "Armand Bashige"
date: "2026-04-25"
categories: [data visualization, SNAP, poverty]
---

```{r}
#| echo: false
#| message: false
#| warning: false

# Load required libraries for data wrangling and visualization
library(tidyverse)
library(ggplot2)
```

## Overview

This is a reworked version of my earlier SNAP visualization. In the previous post,
I looked at household size and children share across poverty quartiles — but those
variables turned out not to be significant predictors of enrollment gaps. This
visualization corrects that by asking a better question:

**Which census tracts are enrolling fewer people than their poverty level predicts,
and does this problem exist across all poverty levels or only in the poorest tracts?**

---

## Visualization

```{r}
#| echo: false
#| message: false
#| warning: false

### Load and prepare raw data ------------------------------------------------

# Load the Hennepin County SNAP dataset (tract-level monthly estimates)
HennepinData <- read_csv("../../../../hennepin-county-data-project-2026-tai/data_private/HennepinCounty_SNAP_MFIP_Estimates_Tract_RevA.csv")

### Clean and aggregate -------------------------------------------------------

# Filter to 2023-2025 and average across months per tract
# SNAP and MFIP are mutually exclusive programs so we combine them
# to capture total benefit access across both programs
snap_recent <- HennepinData |>
  mutate(
    SNAP_cases     = as.numeric(SNAP_cases),
    SNAP_people    = as.numeric(SNAP_people),
    SNAP_children  = as.numeric(SNAP_children),
    MFIP_cases     = as.numeric(MFIP_cases),
    p_125_povertyE = as.numeric(p_125_povertyE),
    GEOID          = as.character(GEOID),
    NAME           = str_replace_all(NAME, ";", ","),
    Year           = YearMonth %/% 100  # Extract 4-digit year from YYYYMM format
  ) |>
  filter(Year %in% c(2023, 2024, 2025)) |>
  filter(
    !is.na(SNAP_cases),
    !is.na(SNAP_people),
    !is.na(SNAP_children),
    !is.na(MFIP_cases),
    !is.na(p_125_povertyE)
  ) |>
  group_by(GEOID, NAME) |>
  summarise(
    SNAP_cases    = mean(SNAP_cases,     na.rm = TRUE),
    SNAP_people   = mean(SNAP_people,    na.rm = TRUE),
    SNAP_children = mean(SNAP_children,  na.rm = TRUE),
    MFIP_cases    = mean(MFIP_cases,     na.rm = TRUE),
    poverty_pop   = mean(p_125_povertyE, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    # Combined take-up rate capped at 1.5 to handle small tract outliers
    combined_cases = SNAP_cases + MFIP_cases,
    snap_takeup    = pmin(
      combined_cases / ifelse(poverty_pop == 0, NA, poverty_pop),
      1.5
    ),
    # Divide tracts into 4 equal groups based on poverty population
    pov_quartile = ntile(poverty_pop, 4),
    pov_label = case_when(
      pov_quartile == 1 ~ "Q1 Low Poverty",
      pov_quartile == 2 ~ "Q2",
      pov_quartile == 3 ~ "Q3",
      pov_quartile == 4 ~ "Q4 High Poverty"
    ),
    # Set factor order so panels display left to right lowest to highest poverty
    pov_label = factor(
      pov_label,
      levels = c("Q1 Low Poverty", "Q2", "Q3", "Q4 High Poverty")
    )
  ) |>
  filter(!is.na(snap_takeup))

### Flag underenrolling tracts ------------------------------------------------

# Use linear regression to find tracts enrolling below
# what we would expect given their poverty level
lm_model <- lm(SNAP_people ~ poverty_pop, data = snap_recent)

snap_recent <- snap_recent |>
  mutate(
    expected_snap     = predict(lm_model, newdata = snap_recent),
    enrollment_gap    = SNAP_people - expected_snap,
    # Tracts below the regression line are enrolling less than expected
    enrollment_status = case_when(
      SNAP_people < expected_snap ~ "Below Line (Underenrolling)",
      TRUE                        ~ "Above Line (Enrolling as Expected)"
    )
  )

### Build the plot ------------------------------------------------------------

ggplot(snap_recent, aes(
  x     = poverty_pop,
  y     = snap_takeup,
  color = enrollment_status
)) +
  geom_point(alpha = 0.55, size = 1.8) +
  # Log scale on x axis because poverty population is right-skewed
  scale_x_log10(labels = scales::comma) +
  scale_color_manual(values = c(
    "Below Line (Underenrolling)"        = "#D62728",
    "Above Line (Enrolling as Expected)" = "#1F77B4"
  )) +
  # One panel per poverty quartile
  facet_wrap(~ pov_label, scales = "free_x") +
  labs(
    title    = "Where Is the Enrollment Gap? (2023-2025)",
    subtitle = "Red tracts have high poverty but lower SNAP/MFIP take-up than expected",
    caption  = "Underenrollment exists across ALL poverty levels, not only in high-poverty tracts.",
    x        = "People at 125% Poverty Level (log scale)",
    y        = "Combined SNAP/MFIP Take-Up Rate",
    color    = "Enrollment Status"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text      = element_text(face = "bold"),
    plot.title      = element_text(face = "bold"),
    plot.caption    = element_text(color = "gray40", size = 9)
  )
```

## How to Read This Plot

The plot is split into four panels from left to right based on poverty level —
Q1 is the least poor and Q4 is the most poor.

- **X axis** — the number of people living at or below 125% of the poverty line
in that census tract (log scale so small and large tracts are both visible)
- **Y axis** — the combined SNAP and MFIP take-up rate. Higher means more eligible
people are actually enrolling
- **Dot color** — red tracts are enrolling fewer people than a regression model
predicts given their poverty level. Blue tracts are enrolling at or above expectations
- **Panels** — splitting by poverty quartile lets us see whether underenrollment
is only a high-poverty problem or whether it shows up everywhere

## What Changed From the Previous Version

In my earlier post (`01_my-second-code`) I visualized household size and children
share across the same four poverty quartiles. However, those variables were tested
in a regression model and found to be statistically insignificant (p = 0.770 and
p = 0.539 respectively). This reworked version instead uses a regression-based
enrollment gap to correctly identify which tracts are underenrolling relative to
what their poverty level would predict.

## Key Finding

Underenrollment is not just a high-poverty problem — red tracts appear in every
quartile from Q1 to Q4. This means that even in neighborhoods with relatively
low poverty, some tracts are still not enrolling as many people as expected. The
problem is widespread across Hennepin County and cannot be solved by targeting
only the poorest neighborhoods.