---
title: "SNAP Household Size & Children by Poverty Level"
author: "Armand Bashige"
date: "2026-04-03"
categories: [data visualization, SNAP, poverty]
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(ggplot2)
library(dplyr)
```

## Overview

This visualization explores how household size and children share relate to SNAP 
enrollment across Hennepin County neighborhoods in 2024. We split neighborhoods 
into four poverty groups to see how these patterns change as poverty increases.

The key question we are trying to answer is:

**Are larger and poorer families actually enrolling in SNAP at the rates we would expect?**

---

## Visualization

```{r}
#| echo: false
#| message: false
#| warning: false

# Load the Hennepin County SNAP dataset
df <- read.csv("../../../../hennepin-county-data-project-2026-tai/data_private/HennepinCounty_SNAP_MFIP_Estimates_Tract_RevA.csv")

# Convert columns to numeric and extract year
df$SNAP_cases_n    <- as.numeric(df$SNAP_cases)
df$SNAP_people_n   <- as.numeric(df$SNAP_people)
df$SNAP_children_n <- as.numeric(df$SNAP_children)
df$Year            <- df$YearMonth %/% 100

# Aggregate to annual tract averages
annual <- df |>
  group_by(GIS_Tract, Year) |>
  summarise(
    SNAP_cases    = mean(SNAP_cases_n,    na.rm = TRUE),
    SNAP_people   = mean(SNAP_people_n,   na.rm = TRUE),
    SNAP_children = mean(SNAP_children_n, na.rm = TRUE),
    poverty_pop   = mean(p_125_povertyE,  na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(!is.na(SNAP_cases), !is.na(SNAP_people),
         !is.na(SNAP_children), !is.na(poverty_pop))

# Calculate derived variables
annual$snap_hh_size     <- annual$SNAP_people   / ifelse(annual$SNAP_cases  == 0, NA, annual$SNAP_cases)
annual$children_share   <- annual$SNAP_children / ifelse(annual$SNAP_people == 0, NA, annual$SNAP_people)
annual$snap_per_poverty <- annual$SNAP_cases    / ifelse(annual$poverty_pop == 0, NA, annual$poverty_pop)

# Filter to 2024 and build poverty quartile labels
df_2024 <- annual |>
  filter(Year == 2024) |>
  filter(!is.na(snap_hh_size), !is.na(children_share)) |>
  mutate(
    pov_q     = ntile(poverty_pop, 4),
    pov_label = case_when(
      pov_q == 1 ~ "Q1 Low poverty",
      pov_q == 2 ~ "Q2",
      pov_q == 3 ~ "Q3",
      pov_q == 4 ~ "Q4 High poverty"
    ),
    pov_label = factor(pov_label, levels = c("Q1 Low poverty","Q2","Q3","Q4 High poverty")),
    snap_take_up = pmin(snap_per_poverty, 1.5)
  )

# Plot
ggplot(df_2024, aes(x = children_share * 100, y = snap_hh_size,
                    color = snap_take_up, size = SNAP_cases)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 2.27, linetype = "dashed", color = "#555555",
             linewidth = 0.5, alpha = 0.5) +
  scale_color_gradientn(
    colors = c("lightblue", "blue", "orangered", "brown4"),
    name   = "SNAP cases\nper poverty\npopulation",
    limits = c(0, 1.5),
    labels = c("0", "0.5", "1.0", "1.5+")
  ) +
  scale_size_continuous(
    name   = "Monthly\nSNAP cases",
    range  = c(0.8, 6),
    breaks = c(50, 200, 600)
  ) +
  facet_wrap(~pov_label, nrow = 1) +
  labs(
    title = "SNAP Household Size & Children by Poverty Levels (2024)",
    subtitle = "Each point = one tract  |  Color = SNAP take-up rate  |  Dashed line = ACS county avg HH size (2.27)",
    caption = "Sources: Hennepin County SNAP/MFIP Estimates  |  283 tracts  |  Implied HH size = SNAP people ÷ SNAP cases",
    x = "% of SNAP recipients who are children",
    y = "Average household size (people per SNAP case)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title        = element_text(face = "bold", size = 13, color = "midnightblue"),
    plot.subtitle     = element_text(size = 9, color = "gray45"),
    plot.caption      = element_text(size = 7.5, color = "gray65"),
    strip.text        = element_text(face = "bold", size = 9.5),
    legend.title      = element_text(size = 8.5),
    legend.text       = element_text(size = 8),
    legend.key.height = unit(0.6, "cm"),
    panel.grid.minor  = element_blank(),
    plot.background   = element_rect(fill = "white", color = NA)
  )
```

## How to Read This Plot

The plot is split into four panels from left to right based on poverty level — 
Q1 is the least poor and Q4 is the most poor.

- **X axis** — what percentage of SNAP recipients in that neighborhood are children. 
Higher means more kids are getting SNAP
- **Y axis** — average household size. Higher means bigger families are on SNAP
- **Dashed line** — the county average household size of 2.27 people per household
- **Dot color** — the SNAP take-up rate. Light blue means few eligible people are 
enrolling, while orange and red means most eligible people are enrolling
- **Dot size** — how many SNAP cases are in that neighborhood

## Key Finding

As poverty increases from Q1 to Q4, households get larger and have more children. 
However, the dots in Q4 are mostly blue — meaning even the poorest neighborhoods 
with the biggest families are not enrolling at the rates we would expect. This 
suggests that poverty and family size alone do not guarantee SNAP enrollment, and 
there may be other barriers keeping eligible families from accessing the program.